---
title: "Word Bank API and `wbstats` wrapper"
output: html_document
---

### Word Bank database info

* [World Bank](https://www.worldbank.org/): this website has many socioeconomic indicators spanning several decades and dozens of topics. Their data is available for bulk download as CSV files, and you previously practiced importing this data in HW5. However, frequently you only need to obtain a handful of indicators or a subset of countries.

* [World Bank API](https://datahelpdesk.worldbank.org/knowledgebase/topics/125589): use it for querying and obtaining a portion of their database using code. 

* [`wbstats`](https://github.com/gshs-ornl/wbstats): a wrapper for R for easy access to the API and returns data in a tidy format.

```{r}
# if you are working with your local R, install wbstats first
library(wbstats)
library(tidyverse)
```


### The`wbstats` wrapper: search for data with `wb_cachelist` and `wb_search`

Let's see a few examples on how we can interact with this API using the `wbstats` wrapper. 

To develop this code, I followed the wrapper documentation and examples, provided in the wrapper GitHub repository: `https://github.com/gshs-ornl/wbstats/blob/master/vignettes/wbstats.md`

`wb_cachelist`: contains a snapshot of available countries, indicators, and other relevant info obtainable through this API:
```{r wb-cachelist}
# display on.y the top level of the object structure
str(wb_cachelist, max.level = 1)
```

Refining our search:
```{r}
glimpse(wb_cachelist$countries)
glimpse(wb_cachelist$indicators)
```

Refining our search even more:
```{r}
wb_cachelist$indicators
wb_cachelist$indicators$indicator
```

For easier access to the indicators, we can use the [web interface](https://data.worldbank.org/indicator) to determine specific indicators and their IDs. 

For example, let's imagine you only need data about unemployment. In the search bar of the web interface, type your search word (here unemployment). You will see all indicators related to unemployment. Explore a few of them. Notice you need to click on "Details" to get the indicator ID

Then, go back to R and do the same using the function `wb_search()`

```{r}
wb_search("unemployment")
```


### The`wbstats` wrapper: download data with `wb_data()`

Once you have the set of indicators you would like to obtain, you can use the `wb_data()` function to generate the API query and download the results. 

For example, if we want to obtain information on [the percent of females participating in the labor force](https://data.worldbank.org/indicator/SL.TLF.TOTL.FE.ZS?view=chart). The indicator ID is `SL.TLF.TOTL.FE.ZS`. 

We can download the indicator for all countries and from 1990 to 2023:

```{r female-labor}
female_labor <- wb_data(
  indicator = "SL.TLF.TOTL.FE.ZS",
  start_date = 1990,
  end_date = 2023
)
female_labor
```

Note the column containing our indicator uses the indicator ID as its name. This is hard to read, so we can adjust it directly in the function:

```{r female-labor-clean}
female_labor <- wb_data(
  indicator = c("fem_lab_part" = "SL.TLF.TOTL.FE.ZS"),
  start_date = 1990,
  end_date = 2020
)
female_labor
```

NB: **do not run the same APIs query code multiple times!** Run it once, and when you have collected your data, save them in a .`csv` file, and import them back (vs. interactively download them every time you run the script). Like that:

```{r}
# write data to csv
write_csv(female_labor, "wb_female_labor.csv")
```

```{r}
# import data back
f_labor <- read_csv("wb_female_labor.csv")
glimpse(f_labor)
```

### Visualize the data in `ggplot2`

Once the data are obtained, we can analyze them, for example:

```{r female-labor-plot, message = FALSE, warning = FALSE}

ggplot(data = f_labor,
       mapping = aes(x = date, y = fem_lab_part)) +
  geom_line(mapping = aes(group = country), alpha = .1) +
  geom_smooth() +
  # https://rdrr.io/cran/scales/man/percent_format.html
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Labor force participation grouped by country",
    x = "Year",
    y = "Percent of total labor force which is female",
    caption = "Source: World Bank"
  )
```

By default, `wb_data()` returns queries as data frames in a wide format. So if we request multiple indicators from the API, each indicator will be stored in its own column:

```{r message = FALSE, warning = FALSE}
female_vars <- wb_data(
  indicator = c(
    "fem_lab_part" = "SL.TLF.TOTL.FE.ZS",
    "fem_educ_sec" = "SE.SEC.CUAT.UP.FE.ZS"
  ),
  start_date = 1990,
  end_date = 2023
)
```

```{r  message = FALSE, warning = FALSE}
ggplot(data = female_vars, mapping = aes(x = fem_educ_sec, y = fem_lab_part)) +
  geom_point(alpha = .2) +
  geom_smooth() +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(
    title = "Female labor force participation",
    x = "Percent of females 25+ who completed secondary school",
    y = "Percent of total labor force which is female",
    caption = "Source: World Bank"
  )
```