---
title: "API username and password using `geonames` as example"
output: html_document
---

**Main goal of this tutorial: illustrate how to store API key and username using the `geonames` API as example** 


### GeoNames geographical database with `geonames`

GeoNames offers geographical information for all countries and other locations. The `geonames` package provides a wrapper for R.

```{r geonames, message = FALSE}
# install.packages(geonames)
library(geonames)
library(tidyverse)
```


### API authentication

**Many APIs require you to register for access.**

* Sometimes, registration is as simple as providing an email and password, then receiving an email with your username and private API key.
* Other times, you need to submit an application and go through a review process.
* Often, this process is free, but some APIs require paying a fee.

**Why register for access?** 

* Registration allows APIs to track users, their queries, and manage demand. 
* If you submit too many queries too quickly, you might be **rate-limited** and your requests de-prioritized or blocked: when in doubt, check the API access policy of the web site to determine what these limits are.


### Register to `GeoNames` API

Go to [the geonames site](http://www.geonames.org/login/) and register an account 

Click [here](http://www.geonames.org/enablefreewebservice) to enable the free web service


### Tell R your `GeoNames` username and key

**To provide R with your API username (and key, if necessary), you have two options** (applicable to all APIs):

* Direct Input: Enter them directly in the R script you are working on.
* Secure Storage: Save them in a secured location, typically in the `.Rprofile` file.

Type this code in R, before interacting with the API: 
```{r}
# replace "my_user_name" with your Geonames username
#options(geonamesUsername = "my_user_name") 

# same for the password
#options(geonamesKey = "your_password")

# then you type code to interact with the api

options(geonamesUsername = "nardin")
options(geonamesKey = "UWBTwhtMuv")
```

This method is OK for a quick check but it is insecure! Especially if you put your work on GitHub. You don't want to commit this line and push it to our public GitHub page

Instead, you should create a file in the same place as your `.Rproj` like this:

  * open the R Console and run: `usethis::edit_r_profile(scope = "project")`. This will create a special file called `.Rprofile` in the same directory as your `.Rproj` file (this assumes you are working in an R project)
  * the file should open automatically; if not, open the file manually and add the same code above to it
      * ensure your `.Rprofile` file ends with a blank line
      * if you push this on GitHub, ensure the line `.Rprofile` is included in your `.gitignore` file, otherwise it will be synced with Github
* Restart RStudio after modifying `.Rprofile` in order to load any new keys into memory

Later, in the R script you are working, type the following code to retrieve the info you saved in your `.Rprofile`. Notice the code below might be different depending on the API:

```{r rprofile, eval = FALSE}

# the names in "" should match those you used in your .Rprofile
username <- getOption("geonamesUsername") 
key <- getOption("geonamesKey")
```

This is a way to keep your keys private, especially if you are sharing the same authentication across several projects. Remember that using `.Rprofile` makes your code un-reproducible. In this case, that is exactly what we want!


### Using GeoNames

**What can we do with this API?** Get access to lots of geographical information such as administrative areas, populated places, weather data, etc.

**How can we use the API in R?** 
* [Use the R wrapper package](chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://cran.r-project.org/web/packages/geonames/geonames.pdf)
* [Check the info provided by GeoNames](http://www.geonames.org/export/)
* Remember to pass your personal username and key first, otherwise the API won't work (see above).

For example:

We can use the `GNcountryInfo()` function to get the info recorded for each country:

```{r geonames-country-info}
country_info <- GNcountryInfo()
```

```{r geonames-str}
country_info %>%
  as_tibble() %>%
  glimpse()
```

We can explore another function in this wrapper, `GNsearch()`, to find places called Lancaster and geocode the result (this code comes from the pdf documentation linked above):
```{r}
lanc_df <- GNsearch(name = "Lancaster", country = "UK")
lanc_df

lanc_coords <- lanc_df[1, c("lng", "lat")]
```


