---
title: "HW04: Tidy data and Programming Exercises - Q3 Review"
author: "Sabrina Nardin"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load library and data
```{r, message = FALSE}
library(tidyverse)
data(iris)
#head(iris)
```

### Question 3. Use the iris dataset to explore subsetting with `[]` and `[[ ]]` in R. The dataset contains 150 observations for three species (setosa, virginica, versicolor) of iris flower, and four features measured for each sample. All measurements are given in centimeters.

There are different ways to interpret this question. The key point to remember is that base R and the tidyverse (dplyr) require different code. This tutorial provides a few examples to address the question.

```{r}
# [] df
df <- iris["Sepal.Length"]
class(df)
is.data.frame(df)

# [[]] vector
v <- iris[["Sepal.Length"]]
class(v)
is.vector(v)
```

In this code:

`[]` gives back a dataframe, because `[]` is designed to preserve the structure of the object you are subsetting, including keeping it as a data frame. 
`[]` is used to select rows and columns in a data frame and keeps it as a dataframe

`[[]]` gives back a vector because `[[]]` is designed to access the element inside a data frame or list and returns it as the actual object (a vector, in this case) not wrapped in a data frame or list.
`[[]]` is used to extract elements from lists; we use it here because a dataframe as a whole can be seen as a list because contains multiple elements of different data types, but all of them are of the same length. The individual columns of a dataframe are all vectors, so when you use `[[]]` you get back the single column as a vector 

Now let's subset with a condition: subset the variable or column `Sepal.Length` if its values or rows are bigger than 6

```{r, error = TRUE}
# [] returns a df using filter() from dplyr
df <- iris[1] %>%
  filter(Sepal.Length > 6) 
class(df)
typeof(df)

# [[]] returns a vector but you get an error if you use filter() from dplyr
# because the function filter() from dplyr only works on dataframes!
v <- iris[[1]] %>%
  filter(Sepal.Length > 6) 

# so to use [[]] you need to base R instead
# take a vector first, then apply condition > 6
v <- iris[[1]]
is.vector(v)
v[v > 6]

# or do the same in one step
iris[[1]][iris[[1]] > 6]

# or like this with $ and [dataframe$column,row] which is also base R syntax
iris[iris$"Sepal.Length" > 6, "Sepal.Length"]
iris[iris$Sepal.Length > 6, "Sepal.Length"]
```

When to use "" and when not use them to refer to variable names? It depends. The code below provides examples for three scenarios: only using base R, mixing base R with tidyverse, only using tidyverse 

```{r, error = TRUE}
# IF YOU ARE ONLY USING BASE R

# with "" and base R only: works
iris[["Sepal.Length"]][iris[["Sepal.Length"]] > 6]

# without "" and base R only: does not work 
iris[[Sepal.Length]][iris[[Sepal.Length]] > 6]

# with or without "" and base R only: both options below works because the $ operator accepts both
iris[iris$"Sepal.Length" > 6, "Sepal.Length"]
iris[iris$Sepal.Length > 6, "Sepal.Length"]
```

```{r, error = TRUE}
# IF YOU ARE MIXING BASE R and TIDYVERSE (DPLYR)

# with "" for base R and without "" for dplyr: works
iris["Sepal.Length"] %>%
  filter(Sepal.Length > 6) 

# without "": does not work, because base R wants ""
iris[Sepal.Length] %>%
  filter(Sepal.Length > 6)
```

```{r}
# IF YOU ARE ONLY USING ONLY DPLYR 

# no need to "" and no need to [] or [[]] use only dplyr verbs
iris %>%
  select(Sepal.Length) %>%
  filter(Sepal.Length > 6)
```

Conclusion. It depends on whether you are using base R or the tidyverse (likely dplyr verbs) for subsetting columns, but as a general rule:

* if you use base R for column subsetting: use "" to refer to the column name 
* if you use dplyr for column subsetting: do not use "" to refer to the column name